Experience in Modeling a Microcontroller Instruction Set Using B

### 2012-07-08 papers/ArticleZ80/source/paper_full_fixed.tex
Apliquei os ajustes sugeridos na versão Full.

%TODO: Verificar possíveis diferenças do FORMATO LNCS

@@

This work presents a formal model of the instruction set of the Z80
microcontroller~\cite{Z80_manual}. This microcontroller was chosen
because of several factors: it contains essential and common concepts
of microcontrollers and even microprocessors; it has extensive
available documentation; and it has been widely used.  A first,
partial, Z80 model presented in \cite{Valerio_SBMF09} was fully
verified, but it used constructs that made it impossible to animate 
with existing technology. %%DD

The new Z80 model supports animation in ProB~\cite{proB}.  Essentially, the
new model\footnote{The interested reader in more details is invited to visit
our repository at: http://code.google.com/p/b2asm/.} no longer contains
definitions derived from infinite sets. Using B constructs for modularity,
auxiliary libraries of basic modules were developed as part of the construction
of the microcontroller model. Such a library has many definitions of 
concepts common to microcontrollers. Besides the new Z80 model, a petroleum
production test system was developed using the Z80 assembly language to
analyse the code simulation and the verification process.

@@

%%DD As primeiras frases do parágrafo abaixo não faziam sentido!
The definitions, functions and properties from hardware library and
types library may be reused to model different instruction set
architectures. Nearly 50\% of all these proof obligations shown in
Table~\ref{tab:types} are verified automatically or with few user
pass\footnote{The user pass allows to filter proof tactics that aid
  the prover to find the proof, this is a feature provided by Atelier
  B.}. A large part of the proof obligations require long sessions
with the interactive provers, mostly because they contain complex
arithmetic expressions that the automatic provers in AtelierB have
difficulty handling. However, all these 531 proof obligations were
eventually verified.

\subsection{Verification of hardware library and types library} 
\label{sec:VerificationHardwareLibrary}

This section presents the verification of the aforementioned modules
using AtelierB provers.  By default, AtelierB~\cite{atelierB} does not
generate proof obligations to check the type of functions defined in
the $PROPERTIES$ clause of a machine. For example, the definition $inc
\in \{0,1\} \fun \{0,1\} \land inc = \lambda(x).( x \in \{0,1\} | x +
1)$, does not create a proof obligation to check the consistency of
the definition. However, this definition is inconsistent since
$inc(1)=2$ and $ 2 \not\in \{0,1\}$.

@@

The conversion functions need to satisfy two properties: bijectivity
and equality to the reverse of the dual conversion function. Three
such proof obligations were verified automatically with AtelierB: (\it
uchar\_schar\rm \ $\in$ \textit{UCHAR} $\bij$ \textit{SCHAR}), (\it
schar\_uchar\rm \ $\in$ \textit{SCHAR} $\bij$ \textit{UCHAR}) and (\it
byte\_bv16\rm \ $=$ \it bv16\_byte\rm $^{-1}$).
%\begin{enumerate}
%
%\item (\it uchar\_schar\rm \ $\in$ \textit{UCHAR} $\bij$ \textit{SCHAR})
%\item (\it schar\_uchar\rm \ $\in$ \textit{SCHAR} $\bij$ \textit{UCHAR})
%\item (\it byte\_bv16\rm \ $=$ \it bv16\_byte\rm $^{-1}$) 
%
%\end{enumerate}
For the remaining proof obligations, we again had to create new proof
rules. There are three kinds of proof rules, that we illustrate by giving
one example each:

@@

Eight proof obligations require that conversion functions are finite. %: \it
% byte\_uchar $\in$ $\finset$\rm (\it byte\_uchar\rm ); \it uchar\_byte  $\in$
% $\finset$\rm (\it uchar\_byte\rm ); \it ushort\_bv16 $ \in \finset$\rm (\it
% ushort\_bv16\rm ); \it bv16\_ushort $ \in \finset$\rm (\it bv16\_ushort\rm ); \it
% byte\_schar $ \in \finset$\rm (\it byte\_schar\rm ); \it schar\_byte $ \in
% \finset$\rm (\it schar\_byte\rm ); \it sshort\_bv16 $ \in \finset$\rm (\it
% sshort\_bv16\rm ); and \it bv16\_sshort $ \in \finset$\rm (\it bv16\_sshort\rm).
 These proof obligations were verified using the expression evaluator of ProB.
Finally, we have included rules related to binary arithmetic as developed
in~\cite{Leibnizens}. These rules are used in the interactive theorem prover with
binary arithmetic to prove important properties~\cite{James2010,Dasgupta2006}.

@@

The \textit{Z80} is a CISC microcontroller developed by
\textit{Zilog}~\cite{Z80_manual}. The Z80 is composed by different
elements and a simplified internal organization is shown in the Figure
\ref{fig:DiagramBlock}.  This figure has some important elements of
Z80 CPU: ALU, registers of 8 and 16 bits and input/output ports.

@@

The Z80 uses a unique memory for storing program instructions, data
stack and general-purpose data. The memory has 16-bit addresses and
each address holds a byte.  Thus, the memory is very simple: \it mem
$\in$ \it BV16 $\fun$ \it BYTE\rm.

Instructions can access all the memory addresses and open the door to
the programmer mistakenly access and change data in memory. For added
security, it is desirable that the program instructions has limited
access by region. Thus the designer can specify address regions in a
refinement model to restrict the access from instructions. The address
regions can be specified using constants
($\textit{PROGRAM\_R\_ADR,DATA\_R\_ADR,STACK\_R\_ADR}$), these define
respectively a range of restricted address for: programs instructions,
general purpose data and data stack.
%
%$
%\begin{array}{l}
%\textit{PROGRAM\_R\_ADR} = 0..16384 \land\\
%\textit{DATA\_R\_ADR} = 16385..49151 \land\\
%\textit{STACK\_R\_ADR} = 49152..65535
%\end{array}
%$

@@

\subsection{Modeling the actions and instructions}
%TODO: Fiz uma breve alteração
Each instruction is represented by a B operation in Z80 module. The
main module (\textit{Z80}) has two categories of operations: a
category represents the instructions of microcontrollers and a
category represents the external actions, e.g.  interruptions
actions. All 158 instructions and interruptions were modeled, but only
two important operations are presented here.

@@

\subsection{Verification and validation}% ESTOU TRABALHANDO AQUI
% OK -> como as regras foram utilizadas ?  Respondido na seção das regras
% TODO:ENGLISH REVIEW parágrafo inicial: VISÃO GERAL

@@

%->Explicaçao de mais detalhes da tabela.

@@

The second strategy, called \emph{user pass}, takes advantage of
another feature of the AtelierB prover. We defined a number of proof
scripts which are then applied to all remaining proof obligations. Few
generic user passes are enough to prove many proof obligations. Also,
as the models of many instructionsa are similar, proofs found in
interactive mode may similarly be replayed and discharge other proof
obligations. We have defined a set of 17 user passes that are able
to discharge more than 2500 WD proof obliagations.

@@

There exists several approaches to model hardware and instruction set
using B, not to mention other formal modelling approaches.  The paper
\cite{springerlink:Yuan2011} reports a method to specify, design and
construct sound and complete instruction set model by stepwise
refinement and formal proof using Event-B. It also discusses desirable
properties of an instruction set model. However, our work
\cite{Valerio_SBMF09} and \cite{Subotic2010} use the B method, that
seems quite appropriated to software development, because it has an
implementable language defined, called B0, a model-driven approach
\cite{Dantas_SBMF08} to develop software from the functional
specification level down to assembly, and tools to convert the models
to a programming language.

@@

% Um pequeno resumo do trabalho
This work has shown an approach to the formal modeling of the
instruction set of microcontroller using the B method.  During the
construction of this model, some problems were found in the official
reference for Z80 microcontroller~\cite{Z80_manual}. We have consulted
different sources \cite{Simulator_z80,UndocumentedZ80,Z80_manual} to
avoid such problems in modeling and have developed a case study: a
verified embedded software.  This case study was interesting to as a
feasibility analysis of the approach to derive formally assembly code
from a specification of functional requirements in the B method.

%TODO: COLOCAR AQUI OS PRINCIPAIS BENEFÍCIOS PROPORCIONADOS PELO TRABALHO

% fundamenta melhor o trabalho já desenvolvido. Citar o artigo SBMF e a dissertação. 
% Os meus trabalhos já desenvolvidos/ resultados atuais.
The following works are directly related to the objective this paper.
The approach to develop verified software down to the assembly level
using B was described in~\cite{DantasSemish2008}. A first case study
for this approach was reported in~\cite{Dantas_SBMF08}, presenting
more details as well as a small software that developed up to assembly
level in three different platforms. A short general view of a previous work
were presented in~\cite{Valerio_SBMF09}. The model presented in the 
current paper added support for animation in ProB and the representation
of external actions (interruptions).
%Link com o parágrafo anterior 
This set of papers present an alternative step to software verification
up to assembly language, showing the current difficulties and
suggesting improvements in tools supporting the B method.

@@

% Conclusão: toque final
%Several iterations were needed to provide the good library definitions as well as
%to fine-tune the model of the microcontroller instructions by factoring common
%functionalities into auxiliary definitions.

>>>>

%TODO: Verificar poss√≠veis diferen√ßas do FORMATO LNCS

@@

This work presents a formal model of the instruction set of the Z80
microcontroller~\cite{Z80_manual}. This microcontroller was chosen
because of several factors: it contains essential and common concepts
of microcontrollers and even microprocessors; it has extensive
available documentation; and it has been widely used.  A first,
partial, Z80 model presented in \cite{Valerio_SBMF09} was fully
verified, but it used constructs that made it incapable to animate 
with existing technology. %%DD

The new Z80 model supports animation in ProB~\cite{proB}.  Essentially, the
new model\footnote{The interested reader in more details is invited to visit
our repository at: http://code.google.com/p/b2asm/.} no longer contains
definitions derived from infinite sets and added support to complex
interruptions.Using B constructs for modularity,
auxiliary libraries of basic modules were developed as part of the construction
of the microcontroller model. Such a library has many definitions of 
concepts common to microcontrollers. Besides the new Z80 model, a petroleum
production test system was developed using the Z80 assembly language to
analyse the code simulation and the verification process.

@@

%%DD As primeiras frases do par√°grafo abaixo n√£o faziam sentido!
The definitions, functions and properties from hardware library and
types library may be reused to model different instruction set
architectures. Nearly 50\% of all these proof obligations shown in
Table~\ref{tab:types} are verified automatically or with few user
pass\footnote{The user passes allows to filter proof tactics that aid
  the prover to find the proof, this is a feature provided by Atelier
  B.}. A large part of the proof obligations require long sessions
with the interactive provers, mostly because they contain complex
arithmetic expressions that the automatic provers in AtelierB have
difficulty handling. However, all these 531 proof obligations were
eventually verified.

\subsection{Verification of hardware library and types library} 
\label{sec:VerificationHardwareLibrary}
%candidato para simplificar/cortar: baixa depend√™ncia 
% Coment√°rio do revisor:
%Section 2.6, par 1: it on the edge of what can be understood without previous knowledge of B, which was not described here.  If the problem raised % specific to Atelier B?  If so, why not use another tool?
This section presents the verification of the aforementioned modules
using AtelierB provers.  By default, AtelierB~\cite{atelierB} does not
generate proof obligations to check the type of functions defined in
the $PROPERTIES$ clause of a machine. For example, the definition $inc
\in \{0,1\} \fun \{0,1\} \land inc = \lambda(x).( x \in \{0,1\} | x +
1)$, does not create a proof obligation to check the consistency of
the definition. However, this definition is inconsistent since
$inc(1)=2$ and $ 2 \not\in \{0,1\}$.

@@

The conversion functions need to satisfy two properties: bijectivity
and equality to the reverse of the dual conversion function. Three
such proof obligations were verified automatically with AtelierB: (\it
uchar\_schar\rm \ $\in$ \textit{UCHAR} $\bij$ \textit{SCHAR}), (\it
schar\_uchar\rm \ $\in$ \textit{SCHAR} $\bij$ \textit{UCHAR}) and (\it
byte\_bv16\rm \ $=$ \it bv16\_byte\rm $^{-1}$).
%\begin{enumerate}
%
%\item (\it uchar\_schar\rm \ $\in$ \textit{UCHAR} $\bij$ \textit{SCHAR})
%\item (\it schar\_uchar\rm \ $\in$ \textit{SCHAR} $\bij$ \textit{UCHAR})
%\item (\it byte\_bv16\rm \ $=$ \it bv16\_byte\rm $^{-1}$) 
%
%\end{enumerate}
For the remaining proof obligations, we had to create new proof
rules. There are three kinds of proof rules, that we illustrate by giving
one example each:

@@

%Eight proof obligations require that conversion functions are finite. 
% \it byte\_uchar $\in$ $\finset$\rm (\it byte\_uchar\rm );
% \it uchar\_byte  $\in$ $\finset$\rm (\it uchar\_byte\rm ); 
%\it ushort\_bv16 $ \in \finset$\rm (\it  ushort\_bv16\rm ); 
%\it bv16\_ushort $ \in \finset$\rm (\it bv16\_ushort\rm );
% \it byte\_schar $ \in \finset$\rm (\it byte\_schar\rm ); 
% \it schar\_byte $ \in \finset$\rm (\it schar\_byte\rm );
% \it sshort\_bv16 $ \in \finset$\rm (\it  sshort\_bv16\rm ); 
% and \it bv16\_sshort $ \in \finset$\rm (\it bv16\_sshort\rm).

%%VM: Realizei ajustes p/ responder perguntas do revisor
For the eight remaining proof obligations, we  had to create new proof
rules. The expression evaluator of ProB verified all rules related to functions with 8-bit data,
the rules to functions with 16-bit data were not verified in ProB, because these rules have
extensive space of verification. However, these rules were built according 
to binary system arithmetic developed by~\cite{Leibnizens} and in similar way as functions with 8-bit data.
% These rules were verified using the expression evaluator of ProB.
%Finally, we have included rules related to binary arithmetic as developed
%in~\cite{Leibnizens}. 
These rules are used in the interactive theorem prover with
binary arithmetic to prove basic properties~\cite{James2010,Dasgupta2006}.

@@

The Z80 is a CISC microcontroller developed by \textit{Zilog}~\cite{Z80_manual}. 
The Z80 is composed by different elements and a simplified internal organization is shown in the Figure
\ref{fig:DiagramBlock}.  This figure has some important elements of
Z80 CPU: ALU, registers of 8 and 16 bits and input/output ports.

@@

The Z80 uses a unique memory for storing program instructions, data
stack and general-purpose data. The memory has 16-bit addresses and
each address holds a byte.  Thus, the memory is very simple: \it mem
$\in$ \it BV16 $\fun$ \it BYTE\rm.
 %
By default, the Z80 no limits the access memory. Instructions can access 
all the memory addresses and open the door to the programmer mistakenly access and change data in memory. For added
security, Z80 model supports to limit access by region. Thus the designer can
specify address regions in an instance of Z80 model to restrict the access from instructions. The address
regions can be specified using constants ($\textit{PROGRAM\_R\_ADR,DATA\_R\_ADR,STACK\_R\_ADR}$), these define
respectively a range of restricted address for: programs instructions,
general purpose data and data stack. 
%
%$
%\begin{array}{l}
%\textit{PROGRAM\_R\_ADR} = 0..16384 \land\\
%\textit{DATA\_R\_ADR} = 16385..49151 \land\\
%\textit{STACK\_R\_ADR} = 49152..65535
%\end{array}
%$

@@

\subsection{Modeling the actions and instructions}
%TODO: Fiz uma breve altera√ß√£o
Each instruction is represented by a B operation in Z80 module. The
main module (\textit{Z80}) has two categories of operations: a
category represents the instructions of microcontrollers and a
category represents the external actions, e.g.  interruptions
actions. All 158 instructions and interruptions were modeled, but only
two important operations are presented here.

@@

\subsection{Verification and validation}
% OK -> como as regras foram utilizadas ?

@@

%->Explicação de mais detalhes da tabela.

@@

The second strategy, called \emph{user pass}, takes advantage of
another feature of the AtelierB prover. We defined a number of proof
scripts which are then applied to all remaining proof obligations. Few
generic user passes are enough to prove many proof obligations. Also,
as the models of many instructions are similar, proofs found in
interactive mode may similarly be replayed and discharge other proof
obligations. We have defined a set of 17 user passes that are able
to discharge more than 2500 WD proof obliagations.

@@

There exists several approaches to model hardware and instruction sets
using B, not to mention other formal modelling approaches.  The paper
\cite{springerlink:Yuan2011} reports a method to specify, design and
construct sound and complete instruction set models by stepwise
refinement and formal proof using Event-B. It also discusses desirable
properties of an instruction set model. However, our work
\cite{Valerio_SBMF09} and \cite{Subotic2010} use the B method, that
seems quite appropriated to software development, because it has an
implementable language defined, called B0, a model-driven approach
\cite{Dantas_SBMF08} to develop software from the functional
specification level down to assembly, and tools to convert the models
to a programming language.

@@

% Um pequeno resumo do trabalho
This work has shown an approach to the formal modeling of the
instruction set of microcontroller using the B method.  During the
construction of this model, some problems were found in the official
reference for Z80 microcontroller~\cite{Z80_manual}. We have consulted
different sources \cite{Simulator_z80,UndocumentedZ80,Z80_manual} to
avoid such problems in modeling and have developed a case study: a
verified embedded software.  This case study was interesting too as a
feasibility analysis of the approach to derive formally assembly code
from a specification of functional requirements in the B method.

%TODO: COLOCAR AQUI OS PRINCIPAIS BENEF√çCIOS PROPORCIONADOS PELO TRABALHO

% fundamenta melhor o trabalho j√° desenvolvido. Citar o artigo SBMF e a disserta√ß√£o. 
% Os meus trabalhos j√° desenvolvidos/ resultados atuais.
The following works are directly related to the objective this paper.
The approach to develop verified software down to the assembly level
using B was described in~\cite{DantasSemish2008}. A first case study
for this approach was reported in~\cite{Dantas_SBMF08}, presenting
more details as well as a small software that developed up to assembly
level in three different platforms. A short general view of a previous work
were presented in~\cite{Valerio_SBMF09}. The model presented in the 
current paper added support for animation in ProB and the representation
of external actions (interruptions).
%Link com o par√°grafo anterior 
This set of papers presents an alternative step to software verification
up to assembly language, showing the current difficulties and
suggesting improvements in tools supporting the B method.

@@

% Conclus√£o: toque final
%Several iterations were needed to provide the good library definitions as well as
%to fine-tune the model of the microcontroller instructions by factoring common
%functionalities into auxiliary definitions.
